A note on the instability and degeneracy of deep learning models

### 2017-11-02 writing/paper-note/chapters/instab.Rmd
dr. nordman's changes

We consider the behavior, and the potential impropriety, of sequences of discrete probability models built to incorporate observations of increasing sample size $N$. Interest is in identifying  instability in such models, which is roughly characterized by probabilities with extreme sensitivity to small changes in data configuration.  The concept of instability was introduced in the field of statistical physics (i.e., point processes) by @ruelle1999statistical and then further extended by Schweinberger @schweinberger2011instability for a family of exponential models. At issue, models exhibiting instability are typically undesirable as these tend to provide poor representations of data or data-generation. As an example, such models can include near-degenerate distributions that assign all essential probability mass to only a subset of an overall sample space. The latter issue in connection to degeneracy has been recognized as a concern in that dominant model outcomes  may not resemble observed data [cf. @handcock2003assessing]. As a compounding issue, model instability often has direct negative impacts for statistical inference and computations based on likelihood functions. Namely, volatilities in probability structure can potentially hamper the numerical evaluations required for maximum likelihood estimation as well as other model-based simulations via Markov Chain Monte Carlo (MCMC). These reasons motivate our general study of instability for a broad class of probability models, described next.

@@

For random variables $\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)$ with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function given by
\begin{equation}
(\#eq:expo)
p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
\end{equation}
depending on parameter vector $\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}$ and natural parameter function $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ with fixed positive integers $k$ and $L$ denoting their dimensions. Above, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ is a vector of sufficient statistics, while
$$
\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
$$
denotes the normalizing function with parameter space $\Theta_N$. The natural parameter function $\eta (\boldsymbol \theta)$ has a linear form (i.e., $\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta$ for a given $L \times k$ matrix $\bm{A}$) in many common model formulaitons, though may also be nonlinear (e.g., curved exponential families). In the linear case, $\eta (\boldsymbol \theta) = \boldsymbol \theta$ may be generally assumed in the exponential parameterization with a minor modification to the definition of sufficient statistics $\boldsymbol g_N(\boldsymbol x)$.

@@

For model sequences $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)$ of the exponential type \@ref(eq:expo), such as those in \@ref(eq:mod1)-\@ref(eq:mod2), note that the dimension $k$ of the parameter $\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k$ necessarily remains the same for all sample sizes $N \geq 1$ as the form of the natural parameter function $\eta(\cdot)$ in \@ref(eq:expo) and the number of sufficient statistics $\boldsymbol g_{N}(\boldsymbol x)$ do not depend on $N$. Consequently, $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$. However, this aspect need not be true for other types of FOES models considered in Sections \@ref(restricted-boltzmann-machines) - \@ref(deep-learning), where instead the numbers of parameters and sufficient statistics commonly increase with the sample size $N$.

## Restricted Boltzmann Machines

@@

This definition of S-unstable is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing possibly non-exponential family models (e.g., RBM and DBM models in Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning) as well as a potentially increasing number $q(N)$ of parameters through the parameter sequence $\boldsymbol \theta_N\in \mathbb{R}^{q(N)}$. While this definition differs in form and scope from the original, it does match that in @schweinberger2011instability for the special case of exponential models (cf. Section \@ref(discrete-exponential-family-models) considered there. Section \@ref(illustrations) provides several examples  of unstable models as well as causes for model instability, where the latter may often be traced to issues in model form (i.e., data functions) and/or parameterization. We next describe several potentially undesirable features associated with S-unstable FOES models.

```{remark, label="rmk1", echo=TRUE}
In the definition \@ref(eq:Sun) of S-instability, we note that the numerical measure $\REP(\boldsymbol \theta_N)/N$ of model instability is invariant to *independent  replications* of data. That is, let $M \geq 1$ denote a possible number of replications and consider data $\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)$ formed by $\{ \boldsymbol X^{(j)}_N\}_{j=1}^M$ as $M$  iid replications of a random vector $\boldsymbol X_N=(X_1,\ldots,X_N)$, where the latter follows a FOES model with probabilities \(P_{ \boldsymbol \theta_N}(\boldsymbol x)>0\), $\boldsymbol x\in\mathcal{X}^N$. This leads to a joint model, say $P_{ \boldsymbol \theta_N}(\boldsymbol y)$, $\boldsymbol y\in\mathcal{X}^{NM}$, for $\boldsymbol Y_{N,M}$ consisting of $N*M$ random variables in total. Then, the LREP for $\boldsymbol Y_{N,M}$, scaled by associated size, is given by
\begin{align*}
\frac{1}{NM}\REP_{\boldsymbol Y_{N,M}}( \boldsymbol\theta_N ) &\equiv \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)} \right] \\
&= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ),
\end{align*}
where $\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N )  \equiv \REP( \boldsymbol\theta_N )$ denotes the log-ratio of extremal probabilities for $\boldsymbol X_N$ defined from \@ref(eq:elpr). That is, due to iid properties, the sample-size corrected  LREP for $\boldsymbol Y_{N,M}$ equals the analog, $\REP(\boldsymbol\theta_N )/N$, from the underlying common data model for $\boldsymbol X_N$ alone, regardless of the level $M \geq 1$ of independent replication. Consequently, the definition of an S-unstable  model  is  unaffected by independent replication and all instability properties  may be characterized by those of one observation  from the common FOES model. For computational purposes, this aspect also implies that if the original data $\boldsymbol X_N=(X_1,\ldots,X_N)$ in a FOES model consist of $N$ iid random variables, then the size-scaled log-ratio \@ref(eq:elpr)  may be calculated as
$$
\frac{1}{N}\REP( \boldsymbol\theta_N ) \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) = \log\left[ \frac{\max_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} {\min_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} \right]
$$
based on the extremal probabilities of just one random variable $X_1$.
```

@@

The above model condition incorporates many standard parameterizations and follows, for instance, whenever $P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y)  = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}$ holds for outcomes $\boldsymbol x, \boldsymbol y \in\mathcal{X}^N$ in a FOES model. For instance, this latter condition is fulfilled for all linear exponential families from Section \@ref(discrete-exponential-family-models) (e.g., \@ref(eq:mod1)-\@ref(eq:mod2)) as well as all network models from Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning) (e.g., \@ref(eq:RBM1)-\@ref(eq:RBM2)). When parameters can be tuned in sign with effects prescribed in the model condition PSR, unstable FOES models will exhibit further probability sensitivities, as outlined in the following extension of Theorem \@ref(thm:degenFOES}).

```{corollary, label="sign", echo=TRUE}
Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models satisfying model condition PSR. If the models $P_{\boldsymbol \theta_N}$  are additionally S-unstable, then
\begin{enumerate}[(i)]
\item the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;
\item and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ under $\boldsymbol \theta_N$ from (\ref{eq:mode}), with $0<\epsilon<1$, it holds under $- \boldsymbol \theta_N$ that
    $$ 
    P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \mbox{as $N\to \infty$},
    $$
    while, by Theorem \ref{thm:degenFOES}, $P_{\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N} ) \rightarrow 1$ holds for $\boldsymbol X_N = (X_1,\ldots,X_N)$ under $\boldsymbol \theta_N$.
\end{enumerate}  
```

@@

```{remark, echo=TRUE}
Model condition PSR and Corollary \ref{cor:sign} can be extended to cases where parameter components $\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})$ are not all changed in sign ($-\boldsymbol \theta_N$) but, more generally, are instead altered to another parameter configuration $\boldsymbol \theta_{N}^A = (\boldsymbol \theta_{N,1}^A, \boldsymbol \theta_{N,2}^A)$ involving a switch in sign only among some model parameters $\boldsymbol \theta_{N,2}^A = -\boldsymbol \theta_{N,2}$ with remaining parameters $\boldsymbol \theta_{1,N}^A$ being arbitrarily chosen. If a sign change occurs among parameters ($\pm \boldsymbol \theta_{N,2}$) which dominate the probability structure of the model, then an unstable model under $\boldsymbol \theta_N$ will then imply that many more unstable models exist over a broad spectrum of possibilities for variations $\boldsymbol \theta_{N}^A$ of $\boldsymbol \theta_N$.
```

@@

As a baseline for comparisons, consider a simplistic model for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with uniform probabilities over the sample space, say $P_{\boldsymbol \theta_N}(\boldsymbol x)= |\mathcal{X}|^{-N}$, $\boldsymbol x \in \mathcal{X}^N$, where each random variable has $|\mathcal{X}| \geq 1$ outcomes. In contrast to instability, model probabilities here are completely insensitive to changes in data outcomes across the sample space, and the associated log-ratio of extreme probabilities \@ref(eq:elpr) is
$$
\frac{1}{N} \REP(\boldsymbol \theta_N)=0\quad \;(\mbox{uniform probability model}),
$$ 
which is as small as possible. In fact, a LREP value of zero can only occur for a FOES model having uniform probabilities, and such equi-probability models are always S-stable.

@@

A fundamental model considered in the instability work of @schweinberger2011instability involves a one-parameter exponential model corresponding to \@ref(eq:expo) with a real-valued parameter, say $\boldsymbol \theta_N =  \eta(\boldsymbol \theta_N)\in \mathbb{R}$, and sufficient statistic $\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}$.
For such models,  upon scaling by sample size $N$, the log-ratio of extreme probabilities in \@ref(eq:elpr) for assessing instability becomes
\begin{equation}
(\#eq:UL)
\frac{1}{N}\REP(\boldsymbol \theta_N ) \equiv   |\boldsymbol \theta_N| \frac{(U_N-L_N)}{N} \;\quad \mbox{(one-parameter exponential model)},
\end{equation}
where $U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$  and $L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ denote the maximal and minimal values of the single sufficient statistic. In this case, an S-unstable model results, by definition \@ref(eq:Sun), whenever $\lim_{N\to \infty} |\boldsymbol \theta_N| (U_N-L_N)/N= \infty$ holds or, in other words, if the combined  magnitudes of parameter $|\boldsymbol \theta_N|$ and maximal difference $U_N-L_N$ in statistic values are overwhelmingly large relative to the sample size $N$. If we further assume that $\boldsymbol \theta_N =\theta\in\mathbb{R}\setminus \{0\}$ is a fixed (non-zero) parameter for all $N \geq 1$, as considered in @schweinberger2011instability, then an S-unstable model results solely if the sufficient statistic admits a value $U_N-L_N$ too large relative to number $N$ of observations, i.e., if $(U_N-L_N)/N\to \infty$ as $N\to \infty$.  The latter aspect reflects the definition of @schweinberger2011instability, for this setting, that a real-valued *statistic* $g_N(\boldsymbol x)$ may be classified as *unstable* when  $\lim_{N\to \infty}|(U_N-L_N)/N=\infty$ holds and as *stable* otherwise (e.g., if $\sup_{N \geq 1}(U_N-L_N)/N<\infty$).

For illustration, consider the iid Bernoulli model \@ref(eq:mod1) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with log-odds ratio parameter $\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}$. Remark 1 (Section \@ref(criterion)) then gives the model instability measure \@ref(eq:Sun) directly as
$$
\frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \mbox{(iid Bernoulli model)},
$$
so that an unstable (or stable) model results for a divergent (or bounded) parameter sequence  $|\boldsymbol \theta_N|$. The above instability expression for the Bernoulli model follows as well from  the $N$-scaled LREP value \@ref(eq:UL) for a one-parameter exponential distribution, using that the sufficient statistic involved $g_N(\boldsymbol x)= \sum_{i=1}^N x_i$, $\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N$, has maximum and mimimum values $U_N=N$ and $L_N=0$. In this case, @schweinberger2011instability has noted that the statistic is stable (i.e., bounded $(U_N-L_N)/N=1$) and the Bernoulli model is as well when, in particular, $\boldsymbol \theta_N=\bold \theta \in\mathbb{R}$ is fixed for $N \geq 1$.

>>>>

We consider the behavior, and the potential impropriety, of sequences of discrete probability models built to incorporate observations of increasing sample size $N$. Interest is in identifying  instability in such models, which is roughly characterized by probabilities with extreme sensitivity to small changes in data configuration.  The concept of instability was introduced in the field of statistical physics (i.e., point processes) by @ruelle1999statistical and then further extended by @schweinberger2011instability for a family of exponential models. At issue, models exhibiting instability are typically undesirable as these tend to provide poor representations of data or data-generation. As an example, such models can include near-degenerate distributions that assign all essential probability mass to only a subset of an overall sample space. The latter issue in connection to degeneracy has been recognized as a concern in that dominant model outcomes  may not resemble observed data [cf. @handcock2003assessing]. As a compounding issue, model instability often has direct negative impacts for statistical inference and computations based on likelihood functions. Namely, volatilities in probability structure can potentially hamper the numerical evaluations required for maximum likelihood estimation as well as other model-based simulations via Markov Chain Monte Carlo (MCMC). These reasons motivate our general study of instability for a broad class of probability models, described next.

@@

For random variables $\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)$ with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function given by
\begin{equation}
(\#eq:expo)
p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
\end{equation}
depending on parameter vector $\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}$ and natural parameter function $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ with fixed positive integers $k$ and $L$ denoting their dimensions. Above, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ is a vector of sufficient statistics, while
$$
\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
$$
denotes the normalizing function with parameter space $\Theta_N$. The natural parameter function $\eta (\boldsymbol \theta)$ has a linear form (i.e., $\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta$ for a given $L \times k$ matrix $\bm{A}$) in many common model formulations, though may also be nonlinear (e.g., curved exponential families). In the linear case, $\eta (\boldsymbol \theta) = \boldsymbol \theta$ may be generally assumed in the exponential parameterization with a minor modification to the definition of sufficient statistics $\boldsymbol g_N(\boldsymbol x)$.

@@

For model sequences $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)$ of the exponential type \@ref(eq:expo), such as those in \@ref(eq:mod1)-\@ref(eq:mod2), note that the dimension $k$ of the parameter $\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k$ necessarily remains the same for all sample sizes $N \geq 1$ as the form of the natural parameter function $\eta(\cdot)$ in \@ref(eq:expo) and the number of sufficient statistics $\boldsymbol g_{N}(\boldsymbol x)$ do not depend on $N$. Consequently, $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$. However, this aspect need not be true for other types of FOES models considered in Sections \@ref(rbm) - \@ref(deep-learning), where instead the numbers of parameters and sufficient statistics commonly increase with the sample size $N$.

## Restricted Boltzmann Machines {#rbm}

@@

This definition of S-unstable is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing possibly non-exponential family models (e.g., RBM and DBM models in Sections \@ref(rbm)-\@ref(deep-learning) as well as a potentially increasing number $q(N)$ of parameters through the parameter sequence $\boldsymbol \theta_N\in \mathbb{R}^{q(N)}$. While this definition differs in form and scope from the original, it does match that in @schweinberger2011instability for the special case of exponential models (cf. Section \@ref(discrete-exponential-family-models) considered there. Section \@ref(illustrations) provides several examples  of unstable models as well as causes for model instability, where the latter may often be traced to issues in model form (i.e., data functions) and/or parameterization. We next describe several potentially undesirable features associated with S-unstable FOES models.

```{remark, label="rmk1", echo=TRUE}
In the definition \@ref(eq:Sun) of S-instability, we note that the numerical measure $\REP(\boldsymbol \theta_N)/N$ of model instability is invariant to *independent  replications* of data. That is, let $M \geq 1$ denote a possible number of replications and consider data $\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)$ formed by $\{ \boldsymbol X^{(j)}_N\}_{j=1}^M$ as $M$  iid replications of a random vector $\boldsymbol X_N=(X_1,\ldots,X_N)$, where the latter follows a FOES model with probabilities $P_{ \boldsymbol \theta_N}(\boldsymbol x)>0$, $\boldsymbol x\in\mathcal{X}^N$. This leads to a joint model, say $P_{ \boldsymbol \theta_N}(\boldsymbol y)$, $\boldsymbol y\in\mathcal{X}^{NM}$, for $\boldsymbol Y_{N,M}$ consisting of $N*M$ random variables in total. Then, the LREP for $\boldsymbol Y_{N,M}$, scaled by associated size, is given by
\begin{align*}
\frac{1}{NM}\REP_{\boldsymbol Y_{N,M}}( \boldsymbol\theta_N ) &\equiv \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)} \right] \\
&= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ),
\end{align*}
where $\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N )  \equiv \REP( \boldsymbol\theta_N )$ denotes the log-ratio of extremal probabilities for $\boldsymbol X_N$ defined from \@ref(eq:elpr). That is, due to iid properties, the sample-size corrected  LREP for $\boldsymbol Y_{N,M}$ equals the analog, $\REP(\boldsymbol\theta_N )/N$, from the underlying common data model for $\boldsymbol X_N$ alone, regardless of the level $M \geq 1$ of independent replication. Consequently, the definition of an S-unstable  model  is  unaffected by independent replication and all instability properties  may be characterized by those of one observation  from the common FOES model. For computational purposes, this aspect also implies that if the original data $\boldsymbol X_N=(X_1,\ldots,X_N)$ in a FOES model consist of $N$ iid random variables, then the size-scaled log-ratio \@ref(eq:elpr)  may be calculated as
$$
\frac{1}{N}\REP( \boldsymbol\theta_N ) \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) = \log\left[ \frac{\max_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} {\min_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} \right]
$$
based on the extremal probabilities of just one random variable $X_1$.
```

@@

The above model condition incorporates many standard parameterizations and follows, for instance, whenever $P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y)  = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}$ holds for outcomes $\boldsymbol x, \boldsymbol y \in\mathcal{X}^N$ in a FOES model. For instance, this latter condition is fulfilled for all linear exponential families from Section \@ref(discrete-exponential-family-models) (e.g., \@ref(eq:mod1)-\@ref(eq:mod2)) as well as all network models from Sections \@ref(rbm)-\@ref(deep-learning) (e.g., \@ref(eq:RBM1)-\@ref(eq:RBM2)). When parameters can be tuned in sign with effects prescribed in the model condition PSR, unstable FOES models will exhibit further probability sensitivities, as outlined in the following extension of Theorem \@ref(thm:degenFOES}).

```{corollary, label="sign", echo=TRUE}
Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models satisfying model condition PSR. If the models $P_{\boldsymbol \theta_N}$  are additionally S-unstable, then
\begin{enumerate}[(i)]
\item the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;
\item and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ under $\boldsymbol \theta_N$ from (\ref{eq:mode}), with $0<\epsilon<1$, it holds under $-\boldsymbol \theta_N$ that
    $$ 
    P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \text{as } N\to \infty,
    $$
    while, by Theorem \ref{thm:degenFOES}, $P_{\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N} ) \rightarrow 1$ holds for $\boldsymbol X_N = (X_1,\ldots,X_N)$ under $\boldsymbol \theta_N$.
\end{enumerate}  
```

@@

```{remark, echo=TRUE}
Under the model condition PSR, Corollary \ref{cor:sign} can also be extended to
cases where parameter components $\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})$ (say) are not all changed in sign (e.g., $-\boldsymbol \theta_N$) but, more generally, are instead  altered to another parameter configuration $\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )$ involving a switch in sign only among some dominating model parameters $\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}$ with remaining parameters $\boldsymbol \theta_{1,N}^A$ being arbitrarily chosen. If a change sign occurs among parameters ($\pm \boldsymbol \theta_{N,2}$) which dominate the probability structure of the model, then the results of Corollary  \ref{cor:sign} can still hold  with $\boldsymbol \theta_{N}^A$ replacing $-\boldsymbol \theta_N$; as an example of one sufficient condition, if $\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0$ holds in addition to Corollary~1 assumptions, where
$$
G_N(\boldsymbol x, \boldsymbol \theta) =   \frac{\log P_{\boldsymbol \theta}(\boldsymbol x) - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }{\max_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y)  - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }, \quad \boldsymbol x \in\mathcal{X}^N,
$$
represents a standardized form of $\boldsymbol \theta$-model probabilities, then the results of Corollary \ref{cor:sign} apply to $\boldsymbol \theta_{N}^A$ in addition to $-\boldsymbol \theta_N$. As a consequence, an unstable model under $\boldsymbol \theta_N$ can then imply that many more unstable models exist  over a broader spectrum of possibilities for variations $\boldsymbol \theta_{N}^A$ of $\boldsymbol \theta_N$, which involves some amount of sign change among components of $\boldsymbol \theta_N$. 
```

@@

As a baseline for comparisons, consider a simplistic model for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with uniform probabilities over the sample space, say $P_{\boldsymbol \theta_N}(\boldsymbol x)= |\mathcal{X}|^{-N}$, $\boldsymbol x \in \mathcal{X}^N$, where each random variable has $|\mathcal{X}| \geq 1$ outcomes. In contrast to instability, model probabilities here are completely insensitive to changes in data outcomes across the sample space, and the associated log-ratio of extreme probabilities \@ref(eq:elpr) is
$$
\frac{1}{N} \REP(\boldsymbol \theta_N)=0\quad \;(\text{uniform probability model}),
$$ 
which is as small as possible. In fact, a LREP value of zero can only occur for a FOES model having uniform probabilities, and such equi-probability models are always S-stable.

@@

A fundamental model considered in the instability work of @schweinberger2011instability involves a one-parameter exponential model corresponding to \@ref(eq:expo) with a real-valued parameter, say $\boldsymbol \theta_N =  \eta(\boldsymbol \theta_N)\in \mathbb{R}$, and sufficient statistic $\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}$.
For such models, upon scaling by sample size $N$, the log-ratio of extreme probabilities in \@ref(eq:elpr) for assessing instability becomes
\begin{equation}
(\#eq:UL)
\frac{1}{N}\REP(\boldsymbol \theta_N ) \equiv   |\boldsymbol \theta_N| \frac{(U_N-L_N)}{N} \;\quad \text{(one-parameter exponential model)},
\end{equation}
where $U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$  and $L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ denote the maximal and minimal values of the single sufficient statistic. In this case, an S-unstable model results, by definition \@ref(eq:Sun), whenever $\lim_{N\to \infty} |\boldsymbol \theta_N| (U_N-L_N)/N= \infty$ holds or, in other words, if the combined  magnitudes of parameter $|\boldsymbol \theta_N|$ and maximal difference $U_N-L_N$ in statistic values are overwhelmingly large relative to the sample size $N$. If we further assume that $\boldsymbol \theta_N =\theta\in\mathbb{R}\setminus \{0\}$ is a fixed (non-zero) parameter for all $N \geq 1$, as considered in @schweinberger2011instability, then an S-unstable model results solely if the sufficient statistic admits a value $U_N-L_N$ too large relative to number $N$ of observations, i.e., if $(U_N-L_N)/N\to \infty$ as $N\to \infty$.  The latter aspect reflects the definition of @schweinberger2011instability, for this setting, that a real-valued *statistic* $g_N(\boldsymbol x)$ may be classified as *unstable* when  $\lim_{N\to \infty}|(U_N-L_N)/N=\infty$ holds and as *stable* otherwise (e.g., if $\sup_{N \geq 1}(U_N-L_N)/N<\infty$).

For illustration, consider the iid Bernoulli model \@ref(eq:mod1) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with log-odds ratio parameter $\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}$. Remark 1 (Section \@ref(criterion)) then gives the model instability measure \@ref(eq:Sun) directly as
$$
\frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \text{(iid Bernoulli model)},
$$
so that an unstable (or stable) model results for a divergent (or bounded) parameter sequence  $|\boldsymbol \theta_N|$. The above instability expression for the Bernoulli model follows as well from  the $N$-scaled LREP value \@ref(eq:UL) for a one-parameter exponential distribution, using that the sufficient statistic involved $g_N(\boldsymbol x)= \sum_{i=1}^N x_i$, $\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N$, has maximum and minimum values $U_N=N$ and $L_N=0$. In this case, @schweinberger2011instability has noted that the statistic is stable (i.e., bounded $(U_N-L_N)/N=1$) and the Bernoulli model is as well when, in particular, $\boldsymbol \theta_N=\bold \theta \in\mathbb{R}$ is fixed for $N \geq 1$.

Alternatively, considering a random graph with $N={n \choose 2}$ edges among $n$ nodes, the exponential graph model from \@ref(eq:mod2), when based purely on the number of $g_{2,N}(\boldsymbol x)$ of 2-stars or solely the number $g_{3,N}(\boldsymbol x)$  of triangles, $\boldsymbol x\in\{0,1\}^N$, has an measure of instability from \@ref(eq:Sun) as
$$
\frac{1}{N}\REP(\boldsymbol \theta_N )  = \left\{ \begin{array}{lcl} |\boldsymbol \theta_N| (n-2) && \text{(2-star graph model)}\\
|\boldsymbol \theta_N|(n-2)/3 &&\text{(triangle graph model)},\end{array}\right.
$$
by using the (one-parameter exponential) LREP formula \@ref(eq:UL) with statistic maximums $U_N= N(n-2)$ for 2-stars or $U_N= N(n-2)/3$ for triangles  and with minimums $L_N=0$ in both cases. Because the variable number $N\to \infty$ as the node number $n\to \infty$, both counts of 2-stars and triangles are unstable statistics in the sense of @schweinberger2011instability (i.e., $\lim_{N\to \infty} (U_N-L_N)/N=\infty$). Furthermore, both types of graph models are always S-unstable for all possible of parameter sequences $\boldsymbol \theta_N \in\mathbb{R}$ that are bounded away from zero (i.e.,  $\lim_{N\to \infty}\REP(\boldsymbol \theta_N )/N=\infty$ then  holds, including the fixed parameter case $\boldsymbol \theta_N=\theta\in\mathbb{R}\setminus \{0\}$  from @schweinberger2011instability).
